---
title: "R and tidyverse code checkpoints"
subtitle: "SuffolkEcon"
author: ""
date: 
output: 
  html_notebook:
    number_sections: true
    theme: readable
    highlight: pygments
    toc: true
    toc_float: 
      collapsed: yes
---

 
<a href="https://lrdegeest.github.io/stats/code_checkpoints.nb.html" class="btn btn-warning">**Last updated:** `r Sys.time()` EST </a> 

<a href="#table" class="btn btn-success">**Go to quick tables**</a>
<a href="https://suffolkecon.github.io/" class="btn btn-default">Go home</a>

# Set-up {-}

Load the tidyverse:

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse) 
```

The `tidyverse` provides the data `mpg` to demonstrate methods. `mpg` has the mileage (`cty` for city, `hwy` for highway) for a bunch of different cars:

```{r load data}
data("mpg") # load the data mpg

# view the first 5 rows
mpg %>% 
  slice_head(n=5)
```

# Summarizing Data 

## The pipe and `summarise()`

All summarizing operations work like this: 

```{r, eval=FALSE}
data %>% # the pipe: "and then"
  summarise() 
```


* the pipe (`%>%`, "and then")
* [`summarise()`](https://dplyr.tidyverse.org/reference/summarise.html)

Inside `summarise()` you can use functions like `mean()`, `median()` and so on.


### mean

`mean()`

```{r mean}
mpg %>% # take the data, THEN
  summarise(mean(cty)) # summarize: mean
```

**Note.** `cty` is a numerical variable.

### median

`median()`

```{r median}
mpg %>% # take the data, THEN
  summarise(median(cty)) # summarize: median
```

### standard deviation

`sd()`

```{r sd}
mpg %>% # take the data, THEN
  summarise(sd(cty)) # summarize: standard deviation
```

### counting observations

[`n()` (counting)](https://dplyr.tidyverse.org/reference/n.html)


```{r count}
mpg %>% # take the data, THEN
  summarise(n()) # summarize: count number of observations
```

**Why doesn't `n()` take an argument?** Because it counts the number of rows, not columns.

### multiple summaries

You can put as many functions as you want inside `summarise()`:

```{r summarise}
mpg %>% # take the data, THEN
  summarise(mean(cty), median(cty), sd(cty), n()) # summarize: mean, median, standard deviation, count
```

### naming variables inside `summarise()`

You can name variables inside `summarise()` using standard assignment:

```{r name summarise}
mpg %>% # take the data, THEN
  summarise(avg_cty = mean(cty), median_cty = median(cty)) # summarize: mean, median
```


## Grouping (`group_by()`)  

[`group_by()`](https://dplyr.tidyverse.org/reference/group_by.html):

```{r group_by, message=FALSE}
mpg %>% # take the data, THEN
  group_by(class) %>% # group it by car class (e.g., compact, pickup) THEN
  summarise(mean(cty), median(cty), sd(cty), n()) # summarize
```

**Note.** `class` is a categorical variable.

# Manipulating data

## Selecting columns

`select()` certain columns

```{r no select}
mpg %>% # all the columns
  slice_head(n=5) # view the first 5 rows
```


```{r yes select}
mpg %>% # all the columns
  select(manufacturer, model, hwy) %>% # select some of the columns
  slice_head(n=5) # view the first 5 rows
```

## Slicing

All slice operators begin with `slice_` and take an optional `n=` argument that specifies the number of rows you want to see. More [here](https://dplyr.tidyverse.org/reference/slice.html). 

Subsets with `slice_head()`:

```{r slice_head}
mpg %>% 
  slice_head(n=10) # view the first 10 rows
```


Maximums with `slice_max()`:

```{r slice_max}
mpg %>% 
  slice_max(hwy, n=5) # the five highest highway miles per gallon. if there are ties they all get printed
```

Minimums with `slice_min()`:

```{r slice_min}
mpg %>% 
  slice_min(hwy, n=2) # the two lowest highway miles per gallon. if there are ties they all get printed
```

Random samples with `slice_sample()`:

```{r slice_sample}
mpg %>% 
  slice_sample(n=3) # pick three rows at random
```

### Slicing and grouping

When you `group_by()` a data frame `slice_()` will return subsets of each group:

```{r group_by and slice_}
mpg %>% 
  group_by(class) %>% # group by class 
  slice_min(hwy, n=1) # bottom two hwy by class
```

If a data frame is already grouped then `slice_()` will always subset by group under the hood. If you want subsets not by group then you have to `ungroup()`. See below in the section on `mutate()`. This often happens when you group a data frame to create a new variable. 

## Filtering

[`filter()`](https://dplyr.tidyverse.org/reference/filter.html) with Boolean logic:

* `==`: "equal to"
* `!=`: "not equal to"
* `>`: "greater than"
* `>=`: "greater than or equal to"
* `<`: "less than"
* `<=`: "less than or equal to"
* `&`: "and"
* `|`: "or"

Boolean logic is any test that returns true or false:

```{r boolean equals to}
2 == 3
```

```{r boolean not equals to}
2 != 3
```

```{r boolean less than}
2 < 3
```

Some examples:

```{r filter equals to}
mpg %>% 
  filter(year == "1999") %>% # filter all cars made in 1999
  slice_head(n=5) # view the first 5 rows
```

```{r filter greater than or equals to}
mpg %>% 
  filter(hwy >= 25) %>% # filter all cars with at least 25 mpg highway
  slice_head(n=5) # view the first 5 rows
```


```{r filter many conditions}
mpg %>% 
  filter(year == "1999" & model != "a4" & hwy >= 25) %>% # filter cars made in 1999, that aren't a4's, and have at least 25 mgp highway
  slice_head(n=5) # view the first 5 rows
```


## Mutating

Create and modify variables (data frame columns) with [`mutate()`](https://dplyr.tidyverse.org/reference/mutate.html)

### Create a new variable

For example, create a column called `mean_hwy` that calculates average highway miles per gallon:

```{r mutate dont add column}
mpg %>% 
  mutate(mean_hwy = mean(hwy)) %>% 
  select(manufacturer, model, hwy, mean_hwy) %>% 
  slice_head(n=5)
```

### Create a new variable and store it as a column

To save the new variable you have to re-assign the data frame:

```{r mutate yes add column}
mpg = mpg %>% 
  mutate(mean_hwy = mean(hwy)) 
```

Now `mpg` has a new column called `mean_hwy`.

### Mutating while grouping

When you create a new variable and save it to the data you re-assign the data. 

When grouping with `group_by()` to create a new column, the re-assigned data will be grouped, which can affect the `slice_()` functions. 

The solution is to `ungroup()` after `mutate()`. 

For instance, add a column called "mean_hwy_class" that calculates average highway miles per gallon by class:

```{r group_by and mutate, message=FALSE}
# first create the variable
mpg = mpg %>% # take the data, THEN
  group_by(class) %>% # group by class, THEN
  mutate(mean_hwy_class = mean(hwy)) %>% # create the new variable, THEN
  ungroup() # ungroup the data
```

Why `ungroup()`? Because otherwise `slice_()` and other functions won't return the output you expect. So better safe than sorry. When grouping *to create a variable* make sure you ungroup at the end.

Now we can slice in general:

```{r view group_by mutate output}
mpg %>% 
  select(manufacturer, model, hwy, mean_hwy_class) %>% 
  slice_head(n=2)
```


### Mutating and frequency tables

Use `n()` to calculate frequency and then use `mutate()` to calculate relative frequency:

```{r frequency and relative frequency, message=FALSE}
mpg %>% 
  group_by(class) %>% 
  summarise(frequency = n()) %>% 
  mutate(relative_frequency = frequency / sum(frequency))
```

# Visualizing data

## Plotting distributions 

### Histograms

Histograms with `ggplot` and [`geom_histogram()`](https://ggplot2.tidyverse.org/reference/geom_histogram.html):

```{r histogram}
mpg %>% 
  ggplot(aes(x = cty)) + # blank canvas: choose the data and the x-axis variable
  geom_histogram() # add geom layer: distribution
```

Faceted histograms with [`facet_wrap()`](https://ggplot2.tidyverse.org/reference/facet_wrap.html):

```{r facetted histogram}
mpg %>% 
  ggplot(aes(x = cty)) + # blank canvas: choose the data and the x-axis variable
  geom_histogram() + # add geom layer:  distribution
  facet_wrap(~class) # add facetting (note the "~" before the categorical variable)
```

### Cumulative distributions

Percentiles and cumulative distributions with `ggplot` and [`stat_ecdf()`](https://ggplot2.tidyverse.org/reference/stat_ecdf.html):

```{r percentile (cumulative distribution)}
mpg %>% 
  ggplot(aes(x=cty)) +  # blank canvas: choose the data and the x-axis variable
  stat_ecdf() # add geom layer: cumulative distribution
```

You can add a vertical line with `geom_vline()` to make it easier to see a percentile. For instance, view percent of cars with less than 15 miles per gallon:

```{r percentile (cumulative distribution) with vline}
mpg %>% 
  ggplot(aes(x=cty)) +  # blank canvas: choose the data and the x-axis variable
  stat_ecdf() + # add geom layer: cumulative distribution
  geom_vline(xintercept = 15, color="red") # verticle 
```

### Box plots

[`geom_boxplot()`](https://ggplot2.tidyverse.org/reference/geom_boxplot.html)

```{r box plot x continuous}
mpg %>% 
  ggplot(aes(x = hwy, y = class)) + # x is continuous, y is categorical
  geom_boxplot() 
```

or: 

```{r box plot x categorical}
mpg %>% 
  ggplot(aes(x = class, y = hwy)) + # x is categorical, y is continuous
  geom_boxplot() 
```


## Scatterplot

`geom_point()`:

```{r geom_point}
mpg %>% 
  ggplot(aes(x = cty, y = hwy)) + 
  geom_point() 
```

### Scatterplot with regression line

[`geom_smooth(method = "lm")`](https://ggplot2.tidyverse.org/reference/geom_smooth.html):

```{r geom_smooth}
mpg %>% 
  ggplot(aes(x = cty, y = hwy)) + 
  geom_point() + # scatter-plot
  geom_smooth(method = "lm") # regression line
```

# Correlation and regression

## Covariance

`cov()`:

```{r cov}
mpg %>% 
  summarise(cov(x=cty, y=hwy))
```

## Correlation

`cor()`:

```{r cor}
mpg %>% 
  summarise(cor(x=cty, y=hwy))
```


## Regression

`lm()`

Simple linear regression (one $x$ variable):

$$
\begin{aligned}
y &= f(x) \\
  &= \beta_0 + \beta_1(x) + \epsilon
\end{aligned}
$$
In `lm()` the $y$ and $x$ variables are separated by `~`:

```{r lm1}
mpg %>% 
  lm(formula = cty ~ hwy, data = .) # the "." means "use the data from the pipe %>%"
```

Multiple linear regression (multiple $x$ variables):

$$
\begin{aligned}
y &= f(\mathbf{X}) \\
  &= f(x_1, x_2, \dots) \\
  &= \beta_0 + \beta_1(x_1) + \beta_2(x_2) + \dots + \epsilon
\end{aligned}
$$
In `lm()` the $x$ variables are separated by `+`:

```{r lm2}
mpg %>% 
  lm(formula = cty ~ hwy + cyl + displ, data = .)
```

## `summary()`

View hypothesis tests and regression diagnostics with `summary()`:

```{r summary2}
mpg %>% 
  lm(formula = cty ~ hwy + cyl + displ, data = .) %>% 
  summary()
```

## Plotting regression results

Use the package `sJplot`. (If you forgot how to install package, see [here](https://lrdegeest.github.io/stats/r.html#3_Install_packages).)

```{r library sjplot, message=FALSE}
library(sjPlot)
```

The function you want is `plot_model`. The baseline plot shows the coefficients and standard errors:

```{r sjPlot demo coefficients}
mpg %>% 
  lm(formula = cty ~ hwy + class, data = .) %>% 
  plot_model(model = .)
```

To plot predicted values you can set `type = "pred"` and then choose which coefficient to plot by setting `terms`:

```{r sjPlot demo predicted values}
mpg %>% 
  lm(formula = cty ~ hwy + class, data = .) %>% 
  plot_model(model = ., type = "pred", terms = c("hwy"))
```

The `terms` argument accepts multiple terms:

```{r sjPlot demo predicted values multiple terms}
mpg %>% 
  lm(formula = cty ~ hwy + class, data = .) %>% 
  plot_model(model = ., type = "pred", terms = c("hwy", "class"))
```

This is useful when you have an interaction effect (e.g., `hwy*class`):

```{r sjPlot demo predicted values interaction}
mpg %>% 
  lm(formula = cty ~ hwy + class + hwy*class, data = .) %>% 
  plot_model(model = ., type = "pred", terms = c("hwy", "class"))
```

`sJplot` uses `ggplot` so you can add `ggplot` stuff to it, like a different theme and titles:

```{r sjPlot demo predicted values interaction spiffy}
mpg %>% 
  lm(formula = cty ~ hwy + class + hwy*class, data = .) %>% 
  plot_model(model = ., type = "pred", terms = c("hwy", "class")) + 
  labs(x = "Highway miles per gallon", y = "Predicted city miles per gallon",
       title = "A very interesting linear model", subtitle = "So interesting") + 
  theme_minimal()
```

The package has tons of great features. Check out the [website!](https://strengejacke.github.io/sjPlot/)

# Quick tables {#table .tabset}

## R

**Function**|**Description**
-----|-----
[<code>%>%</code>](https://magrittr.tidyverse.org/reference/pipe.html)|Pipe operator ("and then")
[<code>summarise()</code>](https://dplyr.tidyverse.org/reference/summarise.html)|Summarize a vector or multiple vectors from a data frame
[<code>mean()</code>](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/mean)|Calculate the mean of a vector
[<code>median()</code>](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/median)|Calculate the median of a vector
[<code>sd()</code>](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/sd)|Calculate the standard deviation of a vector
[<code>n()</code>](https://dplyr.tidyverse.org/reference/n.html)|Count the number of observations. Takes no argument.
[<code>group\_by()</code>](https://dplyr.tidyverse.org/reference/group\_by.html)|Group observations by a categorical variable
[<code>select()</code>](https://dplyr.tidyverse.org/reference/select.html)|Select certain columns
[<code>slice\_()</code>](https://dplyr.tidyverse.org/reference/slice.html)|Slice rows from the data
<code>slice\_head(n=5)</code>|View the head (first five rows) of the data. <code>n = </code> can be any number.
<code>slice\_max(var, n=5)</code>|View the rows with the 5 highest values of column "var"
<code>slice\_min(var, n=5)</code>|View the rows with the 5 lowest values of column "var"
<code>slice\_sample(n=5)</code>|Draw 5 rows at random
[<code>filter()</code>](https://dplyr.tidyverse.org/reference/filter.html)|Filter observations
[<code>mutate()</code>](https://dplyr.tidyverse.org/reference/mutate.html)|Create a new vector
[<code>cov()</code>](https://www.rdocumentation.org/packages/pbdDMAT/versions/0.5-1/topics/covariance)|Calculate the covariation between two variables
[<code>cor()</code>](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/cor)|Calculate the correlation between two variables
[<code>lm()</code>](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/lm)|Estimate a linear regression
[<code>ifelse()</code>](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/ifelse)|Create a vector based on a "True/False" test
[<code>ggplot()</code>](https://ggplot2.tidyverse.org/reference/ggplot.html)|Create a base plot
[<code>geom\_histogram()</code>](https://ggplot2.tidyverse.org/reference/geom\_histogram.html)|Histogram
[<code>stat\_ecdf()</code>](https://ggplot2.tidyverse.org/reference/stat\_ecdf.html)|Cumulative distribution plot
[<code>geom\_boxplot()</code>](https://ggplot2.tidyverse.org/reference/geom\_boxplot.html)|Boxplot
[<code>geom\_point()</code>](https://ggplot2.tidyverse.org/reference/geom\_point.html)|Scatter plot
[<code>geom\_smooth(method = "lm")</code>](https://ggplot2.tidyverse.org/reference/geom\_smooth.html)|Regression line

## Math

**Term**|**Meaning**|**Pronunciation**|**Formula/Example**
:-----:|-----|:-----:|:-----:
$x_i$|data point $i$|"x i" | 
$n$|sample size| | 
$N$|population size| | 
$\bar{x}$|the sample mean|"x bar"|$\frac{\sum_{i=1}^n x_i}{n}$
$\mu$|the population mean|"mu"| 
$s^2$|the sample variance| |$\frac{\sum (x_i - \bar{x})^2}{n-1}$
$\sigma^2$|the population variance| | 
$s$|the sample standard deviation| |$\sqrt{s^2}$
$\sigma$|the population standard deviation|"sigma"| 
$z_i$|z-score for observation $i$| |$z_i = \frac{x_i - \bar{x}}{s}$
$\beta$|regression coefficient|"beta"|$y = \beta_0 + \beta_1 x + \epsilon$
$\hat{\beta}$|estimated value of regression coefficient|"beta hat"| 
$\epsilon$|regression error|"epsilon"| 
$\sum$|summation operator|"sum"|$\sum_{i=1}^2 x_i  = x_1 + x_2$
$s_{xy}$|covariation between two variables $x$ and $y$| |$\frac{\sum (x_i - \bar{x})(y_i - \bar{y})}{n-1} \in [-\infty,\infty]$
$r_{xy}$|correlation between two variables $x$ and $y$| |$\frac{s_{xy}}{s_x s_y} \in [-1,1]$