---
title: "Typesetting Math"
subtitle: "SuffolkEcon"
author: ""
output: 
  html_notebook:
    number_sections: true
    theme: readable
    highlight: pygments
    toc: true
    toc_float: 
      collapsed: yes
---


<a href="https://lrdegeest.github.io/stats/code_checkpoints.nb.html" class="btn btn-warning">**Last updated:** `r Sys.time()` EST </a> 

<a href="https://suffolkecon.github.io/" class="btn btn-default">Go home</a>

# Overview

A great feature of R notebooks is that you can typeset math with a few keystrokes.

This is done through LaTeX, a typesetting language supported by R Markdown. It's a whole kettle of fish. All we need from it are the commands to render math. 

You can learn more about LaTeX math [here](https://bookdown.org/yihui/rmarkdown/markdown-syntax.html#math-expressions) and even more [here](https://www.overleaf.com/learn/latex/mathematical_expressions) (plus a handy list of Greek symbols [here](https://www.overleaf.com/learn/latex/List_of_Greek_letters_and_math_symbols)).

# Inline math

Inline meaning "in the line of the sentence" (or something like that).

In the middle of a sentence just wrap the math bits with `$...$`. 

For example, here is a my amazing formula: `$y = f(x)$` $y = f(x)$


# Formulas

In the lectures you've seen plenty of math formulas. They are always inside a "fence" of dollar signs:

```
$$
% math here
$$
```

For instance, y = f(x) = 2x:

$$
y = f(x) = 2x
$$


## Powers

a raised to the b:

```
$$
a^b
$$
```

If the power is more than one character (e.g., 20 instead of 2) just wrap the power in `{}`:

```
$$
a^{20}
$$
```

## Underscores

a underscore b:

```
$$
a_b
$$
```

multiple characters:

```
a_{bc}
```


## Greek symbols

Everybody knows your math looks better in Greek. 

The most common ones are "betas" for regression coefficients

```
$$
\beta
$$
```

and "epsilons" for the error term:

```
$$
\epsilon
$$
```

For instance, a simple linear regression model with an intercept "beta nought" and a slope "beta one" (you need underscores here):

```
$$
y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \epsilon
$$
```


$$
y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \epsilon
$$

## Fractions

```
\frac{numerator}{denominator}
```

For example:


```
$$
\frac{3x + y^2}{8log(x+3)}
$$
```

$$
\frac{3x + y^2}{8log(x+3)}
$$


We can combine this with the partial derivative symbol `\partial` to write the partial derivative of $f(x,y) = x^2 + y^2$:

```
$$
\frac{\partial f}{\partial x} = 2x = f'_x
$$
```

$$
\frac{\partial f}{\partial x} = 2x = f'_x
$$

# Aligned formulas

This is as complicated as it gets for us, and it's still not that bad.

First, everything has to be in an **aligned** "environment":

```
$$
\begin{aligned}
% math here
\end{aligned}
$$
```

then: 

* use `&` to indicate where you want alignment to happen (e.g., `&=` means "align on the equals sign")
* break each line with `\\`

```
$$
\begin{aligned}
 f(x) + h(x) + j(x) &= x + y \\
 g(x) &= z + b
\end{aligned}
$$
```

$$
\begin{aligned}
 f(x) + h(x) + j(x) + cat(x) &= x + y \\
 g(x) &= z + b
\end{aligned}
$$


# Full example

Let's optimize

$$
f(x) = -x^2 + 2x + 4
$$

and put all the steps in an aligned environment, starting with the first-order condition:


```
$$
\begin{aligned}
 f'_x &= 0 \\
 -2x + 2 &= 0 \\
 -2x &= -2 \\
 x  &= \frac{-2}{-2} \\
 x^* &= 1
\end{aligned}
$$
```


$$
\begin{aligned}
 f'_x &= 0 \\
 -2x + 2 &= 0 \\
 -2x &= -2 \\
 x  &= \frac{-2}{-2} \\
 x^* &= 1
\end{aligned}
$$



OK, let's confirm the results with numerical optimization.

First define the function:

```{r}
f = function(x){
  -x^2 + 2*x + 4
}
```

Then optimize it:

```{r}
optimize(f = f, lower = -100, upper = 100, maximum = TRUE)
```

Verify the solution:

```{r}
f(x = 1)
```









